CUBE: A Memory-lite Cosmological $N$-body Algorithm

### 2017-09-24 ms_caf/ms_caf.tex
revised algorithm section.

\begin{figure}
\centering
  \includegraphics[width=1.0\linewidth]{vdisp.pdf}
 \caption{\tcb{please check units}}
\label{fig.vdisp}
\end{figure}

\subsection{Particle velocity storage}\label{ss.velocity}
Similarly, actual velocity in $d$th dimension $v_d$ is decomposed into an 
averaged velocity field on the same coarse grid $v_c$ and a residual $\Delta v$ 
relative to this field: $v_d=v_c+\Delta v$. $v_c$ is always recorded and 
kept updated, and occupies negligible memory. We then divide velocity space 
$\Delta v$ into uneven bins, and use a $n_\nu$-byte integer to indicate
which $\Delta v$ bin the particle is located.

@@

The essential parameter of the velocity distribution is its variance. On 
non-linear scale, the velocity distribution function is non-Gaussian. 
However, to the first order approximation, we simply assume it as Gaussian 
and characterized by the variance
\begin{equation}\label{eq.vdisp}
	\sigma_v(a,k) = \frac{1}{3}  (a H f D)^2 \int_0^kd^3q\frac{P_{L}(q)}{q^2},
\end{equation}
where $a(z)$ is the scale factor, $H(z)$ the Hubble parameter, $D$ is the 
linear growth factor, $f=d \ln D/d\ln a$, and $P_L(k)$ is the linear power 
spectrum of density contrast at redshift zero.
$\sigma_v(a,k)$ is a function of cosmic evolution $a$ and a smoothing scale 
$k$, or $r$ (see Figure \ref{fig.vdisp}). $\Delta v$
is the velocity dispersion relative to the coarse grid, so we approximate 
its variance as
\begin{equation}\label{eq.vdelta}
	\sigma^2_{\Delta}(a)=\sigma^2_v(a,r_c)-\sigma^2_v(a,r_p), 
\end{equation}
where $r_c$ is the scale of
coarse grid, and $r_p$ is the scale of average particle separation. In each
dimension of 3D velocity field, we use $\sigma^2_{\Delta}(a)/3$ according 
to the equipartition therorem. On different scales, we measure the 
statistics of $v_d$, $v_c$ and $\Delta v$ and find good agreement with the 
above model.

@@

The simulation results are very insensitive if we manually tune the variance of the model $\sigma_\Delta$ within an order of magnitude. However, the method of using uneven bins gets much better results than simply using equal bins between minimum and maximum values $[\min(\Delta v),\max(\Delta v)]$. So, one can safely use a standard $\Lambda$CDM (Cold Dark Matter with a cosmological constant) for slightly different cosmological models, in equation (\ref{eq.vdisp}). In {\tt CUBE}, the velocity conversion takes the formula
\begin{equation}\label{eq.nu}
	\nu_d=\left\lfloor(2^{8n_\nu}-1)\pi^{-1}\tan^{-1}\left(v_d\sqrt{\pi/2\sigma_\Delta^2}\right)\right\rfloor,
\end{equation}
\begin{equation}\label{eq.v}
	v_d=\tan\left(\frac{\pi\nu_d}{2^{8n_\nu}-1}\right)\sqrt{2\sigma_\Delta^2/\pi},
\end{equation}
where $\lfloor\ \rfloor$ is the operator to take the nearest integer. 
Tangent functions are convenient and compute very fast. Compared to error 
functions used in Gaussian case, they take the same variance at $v_d=0$ but 
resolve high velocities relatively better.
Note again that proper choice of conversion formulae and $\sigma_\Delta$ 
only optimizes the velocity space sampling, but does not affect the physics.

@@

\subsubsection{Spacial decomposition}\label{sss.spatial_decomposition}
{\tt CUBE} uses cubic decomposition structures. The global simulation volume is decomposed into ${\tt nn}^3$ cubic sub-volumes, and each of these are assigned to a coarray {\it image}. Inside of an image, the sub-volume is further decomposed into ${\tt nnt}^3$ cubic {\it tiles}. Each tile is surrounded by a {\it buffer} region which is ${\tt ncb}$ coarse cells thick. The buffer is designed for two reasons: (1) computing the fine mesh force, whose cut-off ${\tt nforce\_cutoff}\leq{\tt ncb}$, and (2) collecting all possible particles travelling from a tile's buffer region to its center, {\it physical} region. Thus, a integer-based coarse mesh particle number density ($\rho_c$) is defined as
\begin{equation*}
{\tt integer(1)\ rho\_c(nex,nex,nex,nnt,nnt,nnt)[nn,nn,*]}\\
\end{equation*}
where ${\tt nex}={\tt nt}+2\times{\tt ncb}$ covers the buffer region on both sides, {\tt nnt} is the tile dimemsions, and {\tt nn} is the image {\it codimensions}\footnote{Coarray Fortran concept. Codimensions can do communications between images.}. In the buffer region, phase space coordinates of particles {\tt xp} and {\tt vp} are copied from adjacent tiles and images, when necessary. Figure \ref{fig.tile} shows the spacial decomposition in a 2-dimensional analogy, with ${\tt nn=2}$ and ${\tt nnt=2}$.

The particle position and velocity arrays {\tt xp} and {\tt vp} (equivalent to $\chi_d$ and $\nu_d$ in \ref{ss.position} and \ref{ss.velocity} respectively) are required to be sorted according to the same memory layout of {\tt rho\_c}, such that $n_c$ and thus global positions of particles $x_d$ can be obtained by equation (\ref{eq.x}).

@@

{\tt initialize} and {\tt read\_particles} -- The subroutine {\tt initialize} creates necessary FFT plans and read in configuration files telling the program at which redshifts we need to do checkpoints, halofinds, or stop the simulation. Force kernels are also created or read in here. In {\tt read\_particles}, for each image, we read in all particles in {\it physical} regions of every tile (indicated in Figure \ref{fig.tile}), i.e., their initial positions {\tt xp}, velocities {\tt vp} in the compressed format, and their corresponding, physical part of the number density field {\tt rho\_c}. These are obtained by the initial condition generator (see Appendix A). Because physical regions of tile are {\it complete} and {\it disjoint} in space, particles at this stage are also complete and disjoint. We call it ``disjoint state''. At this stage, {\tt rho\_c}'s buffer regions of each tile are 0, and the elements of {\tt xp} and {\tt vp} beyond physical number can be arbitrary and are not used.

{\tt buffer\_density}, {\tt buffer\_x} and {\tt buffer\_v} -- In order to use the integer based format, {\tt xp} and {\tt vp} must always be ordered, and their number density field {\tt rho\_c} must always be present. In updating arrays of {\tt xp} and {\tt vp} (not simultaneously) of a local tile, a vicinity (buffer) region of the physical region is also needed. First, buffer regions of {\tt rho\_c} is synchronized between tiles and images by subroutine {\tt buffer\_density}. Then, by subroutines {\tt buffer\_x} and {\tt buffer\_v} respectively, {\tt xp} and {\tt vp} are updated to contain common, buffered particles, in an order according to the new, buffered {\tt rho\_c}. We call this stage ``buffered state''. After {\tt particle\_initialize} is done by all images (synchronized by ``{\tt sync all}'', which is equivalent to a {\tt mpi\_barrier}), these three subroutines are called and particles are converted from disjoint state to buffered state.

@@

    program CUBE
      call initialize
      call read_particles
      sync all
      call buffer_density
      call buffer_xp
      call buffer_vp
      do
        call timestep
        call update_xp
sys
        call buffer_xp
        call update_vp
        call buffer_vp
        if(checkpoint_step) then
           call update_xp
           call checkpoint
           if (final_step) exit
           call buffer_density
           call buffer_xp
           call buffer_vp
        endif
      enddo
      call finalize
    end
\end{verbatim}
\caption{Overall structure of {\tt CUBE}.}
\label{fig.code}
\end{figure}

{\tt timestep} --
We operate a Runge-Kutta 2 method for time integration. i.e., we update position ($D$=drift) and velocity ($K$=kick) every half time step. For $n$ time steps, the operation would be $(DKKD)^n$ which is 2nd order accurate. The actual simulation applies varied time steps. In each iteration of the main loop, we firstly call {\tt timestep}, where a increment of time {\tt dt} is controlled by particles' maximum velocities, accelerations, and cosmic expansions.

{\tt update\_xp} --
According to {\tt dt}, subroutine {\tt update\_xp} updates the positions of particles in a ``{\it gather}'' algorithm (in contrast, {\tt CUBEP3M} uses ``scatter'' algorithm) tile by tile. Because each tile is in the buffered state with buffer depth {\tt ncb}, we are able to collect all possible particles whose {\tt vp}*{\tt dt}$<${\tt ncb} from physical+buffer region to its physical region. In order to keep particles ordered, we have to execute ${\tt xp}={\tt xp}+{\tt vp}*{\tt dt}$ twice, first time to obtain an updated density field on the tile, {\tt rho\_c\_new}, and second time to generate a new, local particle list {\tt xp\_new} and {\tt vp\_new} on the tile. The reason for this repeated execution is the dependence of {\tt x\_new} and {\tt v\_new}'s value and ordering on both the old {\tt xp}, {\tt vp} and the new {\tt rho\_c\_new}. However, ${\tt xp}={\tt xp}+{\tt vp}*{\tt dt}$ scales as $o(N)$ and is computational inexpensive. Although {\tt xp\_new} and {\tt vp\_new} arrays take extra memory, they do not appear simultaneously for multiple tiles, and compared to {\tt xp} and {\tt vp}, the extra memory overhead is by a factor of $1/{\tt nnt}^3$. This requires the ``gather'' algorithm to move particles -- as we synchronize the correct \{{\tt rho\_c}, {\tt xp}, {\tt vp}\} in the buffer region of each tile, and the particles in the extended (physical+buffer) region are a superset of particles locating in the physical region in the next time step, given that ${\tt vp}*{\tt dt}$ is not greater than the buffer depth. In {\tt update\_xp}, \{{\tt rho\_c\_new}, {\tt xp\_new}, {\tt vp\_new}\} can be set to collect only the physical region of the tile, with the rest of this subset discarded. At the end, \{{\tt rho\_c}, {\tt xp}, {\tt vp}\} is replaced by \{{\tt xp\_new}, {\tt vp\_new}, {\tt rho\_c\_new}, {\tt vfield}\}, and the memory of latter is freed up. After this step, the tile does not contain buffered regions, being a disjoint state.

{\tt update\_vp} -- 
After drift $D$, we need to update velocities (kick $K$) of particles in physical region of each tile. Before this step, we need to call {\tt buffer\_density} and {\tt buffer\_xp} in order that particle positions are in buffered state. {\tt CUBE} uses a 2-level particle mesh scheme \citep{2013MNRAS.436..540H}. Local fine forces have a force cutoff ${\tt nforce\_cutoff}\leq{\tt ncb}$. So, if we apply a fine-grid particle-mesh on an extended tile with buffered-depth {\tt ncb} (see Figure \ref{fig.tile} regard it periodic), the force on physical regions does not depend on the false periodic boundary assumption, and the resulting fine force and velocity update on physical region is correct.

The compensating coarse grid force is globally computed by using a coarser (usually by factor of 4) mesh by dimensional splitting -- a distributed-memory cubic decomposed 3D coarse density field is interpolated by particles, and we Fourier transform data in consecutive three dimensions with global transposition in between (known as the pencil decomposition). After the multiplication of force kernels, the inverse transform takes place to get the cubic distributed coarse force field, upon which velocities are updated again.

An optional particle-particle (PP) force can be called to increase the force resolution and the velocities are updated last time according to this. \tcb{v field}. During each of these force calculations, maximum accelerations are collected serving for the {\tt timestep} in the next iteration, controlling next {\tt dt}.

After all possible velocity updates, {\tt vp} in physical regions are correctly updated. Particle locations in the buffer regions has updated before {\tt PM} and remained unchanged. So we simply call {\tt buffer\_v} again such that the {\tt update\_x} in the next iteration will be done correctly. We call {\tt update\_vp} to convert {\tt vp} into buffered state.

{\tt checkpoint} -- 
If a desired redshift is reached, we execute the last drift step in the $(DKKD)^n$ operation by {\tt update\_xp}, and call {\tt checkpoint} to save the disjoint state of \{{\tt xp\_new}, {\tt vp\_new}, {\tt rho\_c\_new}, {\tt vfield}\} on disk. Related operations, like run-time halo finder, projections are also done at this point. If final desired redshift is reached, {\tt final\_step} let us exit the main loop. These corresponding logical variables are controlled in {\tt timestep}.

@@

\subsection{Memory layout}\label{ss.memory}
The memory overhead of fine force calculation is controlled by $(1/{\tt nnt})^3$. The memory usage of coarse arrays are negligible.

Here we summarize the memory usage in unit of bpp (byte per particle).

>>>>

\begin{figure}
\centering
  \includegraphics[width=1.0\linewidth]{vdisp.pdf}
 \caption{\tcb{Xin, could you please check the units of both $y$-axis?}}
\label{fig.vdisp}
\end{figure}

\subsection{Particle velocity storage}\label{ss.velocity}
Similarly, actual velocity in $d$th dimension $v_d$ is decomposed into an 
averaged velocity field on the same coarse grid $v_c$ and a residual $\Delta v$ 
relative to this field: 
\begin{equation}
	v_d=v_c+\Delta v.
\end{equation}
$v_c$ is always recorded and 
kept updated, and occupies negligible memory. We then divide velocity space 
$\Delta v$ into uneven bins, and use a $n_\nu$-byte integer to indicate
which $\Delta v$ bin the particle is located.

@@

The essential parameter of the velocity distribution is its variance. On 
non-linear scale, the velocity distribution function is non-Gaussian. 
However, to the first order approximation, we simply assume it as Gaussian 
and characterized by the variance
\begin{equation}\label{eq.vdisp}
	\tcb{ \sigma_v(a,k) = \frac{1}{3}  (a H f D)^2 \int_0^kd^3q\frac{P_{L}(q)}{q^2}},
\end{equation}
where $a(z)$ is the scale factor, $H(z)$ the Hubble parameter, $D$ is the 
linear growth factor, $f=d \ln D/d\ln a$, and $P_L(k)$ is the linear power 
spectrum of density contrast at redshift zero.
$\sigma_v(a,k)$ is a function of cosmic evolution $a$ and a smoothing scale 
$k$, or $r$ (see Figure \ref{fig.vdisp}). $\Delta v$
is the velocity dispersion relative to the coarse grid, so we approximate 
its variance as
\begin{equation}\label{eq.vdelta}
	\sigma^2_{\Delta}(a)=\sigma^2_v(a,r_c)-\sigma^2_v(a,r_p), 
\end{equation}
where $r_c$ is the scale of
coarse grid, and $r_p$ is the scale of average particle separation. In each
dimension of 3D velocity field, we use $\sigma^2_{\Delta}(a)/3$ according 
to the equipartition therorem. On different scales, we measure the 
statistics of $v_d$, $v_c$ and $\Delta v$ and find good agreement with the 
above model.

@@

The simulation results are very insensitive if we manually tune the variance of the model $\sigma_\Delta$ within an order of magnitude. However, in $n_\nu=1$ case, the method of using uneven bins gets much better results than simply using equal bins between minimum and maximum values $[\min(\Delta v),\max(\Delta v)]$. So, one can safely use a standard $\Lambda$CDM (cold dark matter with a cosmological constant as dark energy) for slightly different cosmological models, in equation (\ref{eq.vdisp}). In {\tt CUBE}, the velocity conversion takes the formula
\begin{equation}\label{eq.nu}
	\nu_d=\left\lfloor(2^{8n_\nu}-1)\pi^{-1}\tan^{-1}\left((v_d-v_c)\sqrt{\pi/2\sigma_\Delta^2}\right)\right\rfloor,
\end{equation}
\begin{equation}\label{eq.v}
	v_d=v_c+\tan\left(\frac{\pi\nu_d}{2^{8n_\nu}-1}\right)\sqrt{2\sigma_\Delta^2/\pi},
\end{equation}
where $\lfloor\ \rfloor$ is the operator to take the nearest integer. 
Tangent functions are convenient and compute very fast. Compared to error 
functions used in Gaussian case, they take the same variance at $\nu_d=0$ but 
resolve high velocities relatively better.
Note again that proper choice of conversion formulae and $\sigma_\Delta$ 
only optimizes the velocity space sampling, but does not affect the physics.

@@

\subsubsection{Spacial decomposition}\label{sss.spatial_decomposition}
{\tt CUBE} uses cubic decomposition structures. The global simulation volume is decomposed into ${\tt nn}^3$ cubic sub-volumes, and each of these are assigned to a coarray {\it image}. Inside of an image, the sub-volume is further decomposed into ${\tt nnt}^3$ cubic {\it tiles}, which is a essential unit in the calculations. Each tile is surrounded by a {\it buffer} region which is ${\tt ncb}$ coarse cells thick. The buffer is designed for two reasons: (1) computing the fine mesh force, whose cut-off length ${\tt nforce\_cutoff}\leq{\tt ncb}$, and (2) collecting all possible particles travelling from a tile's buffer region to its center, {\it physical} region. Figure \ref{fig.tile} shows the spacial decomposition in a 2-dimensional analogy, with ${\tt nn=2}$ and ${\tt nnt=2}$. According to this spatial decomposition, and as discussed in the last two subsections, we declare $\{\rho_c,\chi_d,v_c,\nu_d\}$, or in variable expression $\{{\tt rho\_c},{\tt xp},{\tt vfield},{\tt vp}\}$ as follows:
\begin{eqnarray*}
	&&{\tt integer(1)\ rho\_c(nex,nex,nex,nnt,nnt,nnt)[nn,nn,*]}\\
	&&{\tt integer(}n_\chi{\tt )\ xp(npmax)[nn,nn,*]}\\
	&&{\tt real(4)\ vfield(nex,nex,nex,nnt,nnt,nnt)[nn,nn,*]}\\
	&&{\tt integer(}n_\nu{\tt )\ vp(npmax)[nn,nn,*]}
\end{eqnarray*}
where ${\tt nex}={\tt nt}+2\times{\tt ncb}$ covers the buffer region on both sides, {\tt nnt} is the tile dimemsions, and {\tt nn} is the image {\it codimensions}\footnote{Coarray Fortran concept. Codimensions can do communications between images.}. We denote actual number of particles in a given image {\tt nplocal}, and ${\tt npmax}>{\tt nplocal}$ is a value large enough to store buffered particles and should take into account the fact that {\tt nplocal} on some images (with densely clustered structures) could be large. The particle position and velocity arrays {\tt xp} and {\tt vp} (equivalent to $\chi_d$ and $\nu_d$ in \ref{ss.position} and \ref{ss.velocity} respectively) are required to be sorted according to the same memory layout of {\tt rho\_c}, such that $n_c$ and thus global positions of particles $x_d$ can be obtained by equation (\ref{eq.x}). $\{{\tt rho\_c},{\tt xp},{\tt vfield},{\tt vp}\}$ provides a complete information of positions and velocities of particles, and we call it a snapshot, or checkpoint.

@@

{\tt initialize} and {\tt read\_particles} -- The subroutine {\tt initialize} creates necessary FFT plans and read in configuration files telling the program at which redshifts we need to do checkpoints, halofinds, or stop the simulation. Force kernels are also created or read in here. In {\tt read\_particles}, for each image, we read in all particles in {\it physical} regions of every tile (indicated in Figure \ref{fig.tile}), i.e., $\{{\tt rho\_c},{\tt xp},{\tt vfield},{\tt vp}\}$. These are obtained by the initial condition generator (see Appendix A). Because physical regions of tile are {\it complete} and {\it disjoint} in space, particles at this stage are also complete and disjoint. We call it ``disjoint state''. At this stage, {\tt rho\_c}'s buffer regions of each tile are 0, and the elements of {\tt xp} and {\tt vp} beyond physical number -- {\tt xp(nplocal+1:)} and {\tt vp(nplocal+1:)} can be arbitrary and are not used.

{\tt buffer\_density}, {\tt buffer\_x} and {\tt buffer\_v} -- In order to use the integer based format, {\tt xp} and {\tt vp} must always be ordered, and their number density field {\tt rho\_c} must always be present. In updating arrays of {\tt xp} and {\tt vp} (not simultaneously) of a local tile, a vicinity (buffer) region of the physical region is also needed. First, buffer regions of {\tt rho\_c} is synchronized between tiles and images by subroutine {\tt buffer\_density}. Then, by subroutines {\tt buffer\_x} and {\tt buffer\_v} respectively, {\tt xp} and {\tt vp} are updated to contain common, buffered particles, in an order according to the new, buffered {\tt rho\_c}. We call this stage ``buffered state''. After {\tt particle\_initialize} is done by all images (synchronized by a ``{\tt sync all}'', which is equivalent to a {\tt mpi\_barrier}), these three subroutines are called and particles are converted from disjoint state to buffered state.

@@

    program CUBE
      call initialize
      call read_particles
      call buffer_density
      call buffer_xp
      call buffer_vp
      do
        call timestep
        call update_xp
        call buffer_density
        call buffer_xp
        call update_vp
        call buffer_vp
        if(checkpoint_step) then
           call update_xp
           call checkpoint
           if (final_step) exit
           call buffer_density
           call buffer_xp
           call buffer_vp
        endif
      enddo
      call finalize
    end
\end{verbatim}
\caption{Overall structure of {\tt CUBE}.}
\label{fig.code}
\end{figure}

{\tt timestep} --
We operate a Runge-Kutta 2 method for time integration. i.e., we update position ($D$=drift) and velocity ($K$=kick) every half time step. For $n$ time steps, the operation would be $(DKKD)^n$ which is 2nd order accurate. The actual simulation applies varied time steps. In each iteration of the main loop, we firstly call {\tt timestep}, where a increment of time {\tt dt} is controlled by particles' maximum velocities, accelerations, cosmic expansion and any other desired conditions.

{\tt update\_xp} --
According to {\tt dt}, subroutine {\tt update\_xp} updates the positions of particles in a ``{\it gather}'' algorithm (in contrast, {\tt CUBEP3M} uses ``scatter'' algorithm) tile by tile. For each particle, {\tt xp} and {\tt vp} are converted to $x_d$ and $v_d$ by Equations(\ref{eq.x},\ref{eq.v}). Because each tile is in the buffered state with buffer depth {\tt ncb}, we are able to collect all possible particles whose $v_d\times{\tt dt}<{\tt ncb}$ from physical+buffer region to its physical region.

In order to keep particles ordered, for each tile, we first do $x_d=x_d+v_d\times{\tt dt}$ on all particles to obtain an updated density and velocity field on the tile, {\tt rho\_c\_new} and {\tt vfield\_new}. Then, this calculation in done on the same tile again\footnote{This repetition scales as $o(N)$ and is computational inexpensive.} to generate a new, local particle list {\tt xp\_new} and {\tt vp\_new} on the tile by Equations(\ref{eq.chi},\ref{eq.nu}). Here, the ordering of particles depends on {\tt rho\_c\_new}, and {\tt vfield\_new} is used as $v_c$ in Equation(\ref{eq.nu}). Then, another iteration is done on this tile to discard buffer regions of $\{{\tt rho\_c\_new},{\tt xp\_new},{\tt vfield\_new},{\tt vp\_new}\}$, converting buffer state to disjoint state, and it replaces corresponding part of $\{{\tt rho\_c},{\tt xp},{\tt vfield},{\tt vp}\}$. When all tiles are iterated, the entire image is in disjoint state, and {\tt nplocal} is updated. Then all images is synchronized by {\tt sync all} and we sum up {\tt nplocal} over images to check if total number of particles {\tt npglobal} is conserved.

{\tt update\_vp} -- 
After {\tt update\_xp}, or drift $D$, we call {\tt buffer\_density} and {\tt buffer\_xp} in order that particle positions are in buffered state, based on which, we update velocities (kick $K$) of particles in physical region of each tile. {\tt CUBE} uses a 2-level particle mesh scheme \citep{2013MNRAS.436..540H}. Local fine forces have a force cutoff ${\tt nforce\_cutoff}\leq{\tt ncb}$, i.e., $F_{\rm fine}(r>{\tt ncb})=0$. So, if we apply a fine-grid particle-mesh on an extended tile with buffered-depth {\tt ncb} (see Figure \ref{fig.tile} and regard it periodic), the force on physical regions does not depend on the false periodic boundary assumption, and the resulting fine force $F_{\rm fine}$ and velocity update on physical region is correct.

The compensating coarse grid force $F_{\rm coarse}$ is globally computed by using a coarser (usually by factor of 4) mesh by dimensional splitting -- a distributed-memory cubic decomposed 3D coarse density field is interpolated by particles, and we Fourier transform data in consecutive three dimensions with global transposition in between (known as the pencil decomposition). After the multiplication of force kernels, the inverse transform takes place to get the cubic distributed coarse force field $F_{\rm coarse}$, upon which velocities are updated again.

An optional particle-particle (PP) force $F_{\rm pp}$ can be called to increase the force resolution and the velocities are updated again. The collective operations of velocity update is Equation(\ref{eq.v}), $v_d=v_d+F_{\rm total}$ and Equation(\ref{eq.nu}).

The maximum accelerations $\max(\dot{v}_{\rm fine})$, $\max(\dot{v}_{\rm coarse})$, $\max(\dot{v}_{\rm pp})$ and maximum of velocities $\max(v_d)$ are collected controlling {\tt dt} for the {\tt timestep} in the next iteration. We also update $\sigma^2_{\Delta}$ according to the new $v_d-v_c$.

By far, {\tt vp} in physical regions are correctly updated. Remember that particle locations in the buffer regions has updated before {\tt PM} and remained unchanged. So we simply call {\tt buffer\_v} again such that the {\tt update\_x} in the next iteration will be done correctly. We call {\tt update\_vp} to convert {\tt vp} into buffered state.

{\tt checkpoint} -- 
If a desired redshift is reached, we execute the last drift step in the $(DKKD)^n$ operation by {\tt update\_xp}, and call {\tt checkpoint} to save the disjoint state of \{{\tt xp}, {\tt vp}, {\tt rho\_c}, {\tt vfield}\} on disk. Related operations, like run-time halo finder, projections are also done at this point. If final desired redshift is reached, {\tt final\_step} let us exit the main loop. These corresponding logical variables are controlled in {\tt timestep}.

@@

\subsection{Memory layout}\label{ss.memory}
The memory overhead of fine force calculation is controlled by $(1/{\tt nnt})^3$. The memory usage of coarse arrays are negligible. 
Although {\tt xp\_new} and {\tt vp\_new} arrays take extra memory, they do not appear simultaneously for multiple tiles, and compared to {\tt xp} and {\tt vp}, the extra memory overhead is by a factor of $1/{\tt nnt}^3$.
Here we summarize the memory usage in unit of bpp (byte per particle).
